---
title: "Module 5: Actual Work"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
    theme: !expr bslib::bs_theme(version = 5, bootswatch = "cerulean")
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(DBI)
library(RPostgres)
library(bslib)
library(shinychat)
library(ellmer)
library(promises)
library(dplyr)
library(ggplot2)
library(rlang)

# Shared database helper functions (get_con, get_or_create_student,
# get_exercise_id, log_attempt, log_chat)
source("R/db_utils.R")
# Tutor chat context assembly, start_tutor_chat(), the shared
# bs5_theme_dependencies()/wire_tutor_chat()/log_attempt_and_maybe_open_chat()
# wiring, and the quiz_question_ui()/quiz_question_server() and
# exercise_ui()/exercise_server() helpers used below in place of
# learnr::question() and exercise=TRUE/gradethis -- see their
# definitions in R/tutor_utils.R for why. exercise_ui()/exercise_server()
# gained a has_plot_output argument for this module -- see R/tutor_utils.R
# for why a ggplot result needs different handling than every other
# module's plain-value results.
source("R/tutor_utils.R")

# --- Practice data ---------------------------------------------------------
# Same already-cleaned fish survey data from Module 4 (each module is its
# own self-contained deployment, so it's recreated here rather than shared
# across modules). See module_4.Rmd for the raw/messy version and the
# renaming exercise that produces this.
fish_clean <- data.frame(
  fish_id   = 1:10,
  species   = c("Bluegill", "Bluegill", "Largemouth Bass", "Largemouth Bass",
                "Yellow Perch", "Yellow Perch", "Bluegill", "Largemouth Bass",
                "Yellow Perch", "Bluegill"),
  length_mm = c(152, 148, 310, NA, 205, 198, 160, 295, 210, 145),
  weight_g  = c(45, 42, NA, 610, 98, 95, 50, 580, NA, 40),
  lake_site = c("North Lake", "North Lake", "South Lake", "South Lake",
                "North Lake", "South Lake", "North Lake", "South Lake",
                "North Lake", "South Lake"),
  stringsAsFactors = FALSE
)

# exercise_server() evaluates submitted code via eval(parse(text = user_code),
# envir = new.env()) -- see the identical note in module_4.Rmd's setup chunk
# for why this needs to be explicitly placed in .GlobalEnv rather than just
# assigned normally.
assign("fish_clean", fish_clean, envir = globalenv())

# Grading logic for each exercise. A ggplot object's mapping (aes()) is
# stored as quosures -- rlang::as_label() turns e.g. aes(x = length_mm)'s
# stored mapping back into the plain string "length_mm" for comparison.
# log_attempt_and_maybe_open_chat() and wire_tutor_chat() themselves are
# shared across modules -- see R/tutor_utils.R.

mapping_label <- function(mapping, aesthetic) {
  if (is.null(mapping[[aesthetic]])) return(NA_character_)
  rlang::as_label(mapping[[aesthetic]])
}

evaluate_scatter_exercise <- function(user_code, result, stage) {
  uses_geom_point <- grepl("geom_point\\s*\\(", user_code)
  is_ggplot <- inherits(result, "ggplot")
  has_point_layer <- is_ggplot && any(vapply(result$layers, function(l) inherits(l$geom, "GeomPoint"), logical(1)))
  x_ok <- is_ggplot && identical(mapping_label(result$mapping, "x"), "length_mm")
  y_ok <- is_ggplot && identical(mapping_label(result$mapping, "y"), "weight_g")

  passed <- uses_geom_point && is_ggplot && has_point_layer && x_ok && y_ok

  feedback_message <- if (passed) {
    "That's a scatter plot of length vs. weight -- exactly right."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check that aes()'s arguments and geom_point() are complete, valid R."
  } else if (!is_ggplot) {
    "This should produce a ggplot object -- make sure ggplot(fish_clean, aes(...)) is still the start of the expression."
  } else if (!has_point_layer) {
    "This needs a geom_point() layer to actually draw the scatter plot."
  } else if (!x_ok) {
    "The x-axis isn't mapped to length_mm yet -- check aes()'s x argument."
  } else {
    "Close -- the y-axis isn't mapped to weight_g yet, check aes()'s y argument."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_bar_exercise <- function(user_code, result, stage) {
  uses_group_by <- grepl("group_by\\s*\\(", user_code)
  uses_geom_col <- grepl("geom_col\\s*\\(", user_code)
  is_ggplot <- inherits(result, "ggplot")
  has_col_layer <- is_ggplot && any(vapply(result$layers, function(l) inherits(l$geom, "GeomCol"), logical(1)))
  right_groups <- is_ggplot && is.data.frame(result$data) && nrow(result$data) == 3
  x_ok <- is_ggplot && identical(mapping_label(result$mapping, "x"), "species")
  y_ok <- is_ggplot && identical(mapping_label(result$mapping, "y"), "mean_weight")

  passed <- uses_group_by && uses_geom_col && is_ggplot && has_col_layer && right_groups && x_ok && y_ok

  feedback_message <- if (passed) {
    "One bar per species, showing each one's mean weight -- exactly right."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- work through each step of the pipeline, same as Module 4."
  } else if (!uses_group_by || !right_groups) {
    "Check the group_by() step -- there are 3 species in this data, and something else is producing a different number of bars."
  } else if (!uses_geom_col || !has_col_layer) {
    "This needs a geom_col() layer to draw the bars."
  } else if (!x_ok) {
    "The x-axis isn't mapped to species yet -- check aes()'s x argument."
  } else {
    "Close -- the y-axis isn't mapped to mean_weight yet, check aes()'s y argument."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_histogram_exercise <- function(user_code, result, stage) {
  uses_geom_histogram <- grepl("geom_histogram\\s*\\(", user_code)
  is_ggplot <- inherits(result, "ggplot")
  has_hist_layer <- is_ggplot && any(vapply(result$layers, function(l) inherits(l$geom, "GeomBar") || inherits(l$geom, "GeomHistogram") || "GeomBar" %in% class(l$geom), logical(1)))
  x_ok <- is_ggplot && identical(mapping_label(result$mapping, "x"), "length_mm")

  passed <- uses_geom_histogram && is_ggplot && has_hist_layer && x_ok

  feedback_message <- if (passed) {
    "That's a histogram of length_mm -- exactly right."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check that aes() and geom_histogram() are complete, valid R."
  } else if (!is_ggplot) {
    "This should produce a ggplot object -- make sure ggplot(fish_clean, aes(...)) is still the start of the expression."
  } else if (!has_hist_layer) {
    "This needs a geom_histogram() layer to actually draw the histogram."
  } else {
    "Close -- the x-axis isn't mapped to length_mm yet, check aes()'s x argument."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_cumulative_plot_exercise <- function(user_code, result, stage) {
  uses_filter <- grepl("filter\\s*\\(", user_code)
  uses_geom_point <- grepl("geom_point\\s*\\(", user_code)
  uses_labs <- grepl("labs\\s*\\(", user_code)
  is_ggplot <- inherits(result, "ggplot")
  has_point_layer <- is_ggplot && any(vapply(result$layers, function(l) inherits(l$geom, "GeomPoint"), logical(1)))
  right_data <- is_ggplot && is.data.frame(result$data) && nrow(result$data) == 4 &&
    "species" %in% names(result$data) && all(result$data$species == "Bluegill")
  has_title <- is_ggplot && is.character(result$labels$title) && nzchar(trimws(result$labels$title))

  passed <- uses_filter && uses_geom_point && uses_labs && is_ggplot && has_point_layer && right_data && has_title

  feedback_message <- if (passed) {
    "That's the full pipeline -- filtered to just Bluegill, plotted, and titled."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- work through it one step at a time, same as you've done in earlier modules."
  } else if (!uses_filter || !right_data) {
    "Check the filter() step -- this plot should only include the 4 Bluegill rows."
  } else if (!uses_geom_point || !has_point_layer) {
    "This still needs a geom_point() layer to draw the scatter plot."
  } else if (!uses_labs || !has_title) {
    "The plot looks right, but it still needs a title -- add labs(title = \"...\") with any title you like."
  } else {
    "Close -- double check each step of the pipeline."
  }

  list(passed = passed, message = feedback_message)
}
```

```{r, context="server"}
# --- Identity capture -------------------------------------------------
# Same pattern as module_1.Rmd/module_2.Rmd/module_3.Rmd/module_4.Rmd: ask
# for a name on start, store the resulting student_id on session$userData
# so exercise check code (which runs semi-isolated from this chunk) can
# reach it via shiny::getDefaultReactiveDomain().

showModal(modalDialog(
  title = "Welcome back!",
  "Please enter your name to begin this tutorial.",
  textInput("student_name", label = NULL, placeholder = "Your name"),
  footer = actionButton("submit_name", "Start Tutorial"),
  easyClose = FALSE
))

observeEvent(input$submit_name, {
  req(input$student_name)
  tryCatch({
    con <- get_con()
    sid <- get_or_create_student(con, trimws(input$student_name))
    dbDisconnect(con)

    session$userData$student_id <- sid
    removeModal()
  }, error = function(e) {
    showModal(modalDialog(
      title = "Couldn't connect",
      paste("There was a problem saving your name (database error):", conditionMessage(e)),
      footer = actionButton("submit_name", "Try again"),
      easyClose = FALSE
    ))
  })
})
```

```{r, context="server"}
# --- Tutor chat wiring --------------------------------------------------
# Each graded exercise gets its own always-available (collapsed by
# default) tutor chat panel. See wire_tutor_chat() in R/tutor_utils.R for
# the full mechanism (fresh client per open, both directions logged).

wire_tutor_chat("module5_scatter", "tutor_chat_1", "tutor_accordion_1", "Ask the Tutor", input, output, session)
wire_tutor_chat("module5_bar", "tutor_chat_2", "tutor_accordion_2", "Ask the Tutor", input, output, session)
wire_tutor_chat("module5_histogram", "tutor_chat_3", "tutor_accordion_3", "Ask the Tutor", input, output, session)
wire_tutor_chat("module5_cumulative", "tutor_chat_4", "tutor_accordion_4", "Ask the Tutor", input, output, session)
```

```{r, context="server"}
# --- Fix for quiz/exercise feedback outputs getting stuck "recalculating"
# if bound while their topic is hidden -- see unstick_hidden_outputs() in
# R/tutor_utils.R.
unstick_hidden_outputs(output, session)
```

```{r, context="server"}
# --- Quiz question wiring ------------------------------------------------
# See quiz_question_ui()/quiz_question_server() in R/tutor_utils.R.

quiz_question_server(input, output, session, "ggplot_grammar_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_speed = list(correct = FALSE, message = "aes() doesn't affect how fast the code runs -- it's purely about what's shown and how."),
  wrong_load = list(correct = FALSE, message = "aes() doesn't load or import anything -- that's a separate step.")
))

quiz_question_server(input, output, session, "markdown_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_speed = list(correct = FALSE, message = "R Markdown doesn't make code run faster -- it's about presentation, not performance."),
  wrong_password = list(correct = FALSE, message = "R Markdown files aren't for storing credentials -- that's what .Renviron (and never committing it to git) is for.")
))

quiz_question_server(input, output, session, "scripts_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_required = list(correct = FALSE, message = "R scripts run from anywhere -- the issue with the Desktop is organization, not a hard requirement."),
  wrong_speed = list(correct = FALSE, message = "Where a script is saved doesn't affect how fast it runs.")
))

quiz_question_server(input, output, session, "naming_files_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_final = list(correct = FALSE, message = "Names like that (with \"final\", spaces, and version numbers in parentheses) are exactly what to avoid -- they don't say what the file actually does, and stack up confusingly over time."),
  wrong_vague = list(correct = FALSE, message = "A name like that doesn't describe what the file does at all -- six months from now, you (or a teammate) would have no way to tell.")
))

quiz_question_server(input, output, session, "version_control_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_autofix = list(correct = FALSE, message = "Git doesn't fix bugs for you -- it just keeps a history of every change, bugs included."),
  wrong_standalone = list(correct = FALSE, message = "Git is unrelated to whether RStudio is installed -- it's about tracking change history, not packaging software.")
))
```

```{r, context="server"}
# --- Exercise wiring -------------------------------------------------------
# See exercise_ui()/exercise_server() in R/tutor_utils.R. has_plot_output =
# TRUE here so a ggplot result is actually rendered, not printed as text.

exercise_server(
  input, output, session, "scatter_exercise", "module5_scatter", "tutor_accordion_1",
  starting_code = "ggplot(fish_clean, aes(x = FIXME, y = FIXME)) +\n  geom_point()",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_scatter_exercise(user_code, result, stage),
  has_plot_output = TRUE
)

exercise_server(
  input, output, session, "bar_exercise", "module5_bar", "tutor_accordion_2",
  starting_code = "fish_clean |>\n  group_by(FIXME) |>\n  summarize(mean_weight = mean(weight_g, na.rm = TRUE)) |>\n  ggplot(aes(x = FIXME, y = FIXME)) +\n  geom_col()",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_bar_exercise(user_code, result, stage),
  has_plot_output = TRUE
)

exercise_server(
  input, output, session, "histogram_exercise", "module5_histogram", "tutor_accordion_3",
  starting_code = "ggplot(fish_clean, aes(x = FIXME)) +\n  geom_histogram(bins = 8)",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_histogram_exercise(user_code, result, stage),
  has_plot_output = TRUE
)

exercise_server(
  input, output, session, "cumulative_plot_exercise", "module5_cumulative", "tutor_accordion_4",
  starting_code = "fish_clean |>\n  filter(FIXME) |>\n  ggplot(aes(x = length_mm, y = weight_g)) +\n  geom_point() +\n  labs(title = FIXME)",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_cumulative_plot_exercise(user_code, result, stage),
  has_plot_output = TRUE
)
```

## Welcome back

```{r, echo=FALSE}
bs5_theme_dependencies()
```

Module 4 got `fish_clean` -- the same fish survey data you'll use again here -- inspected, cleaned, and summarized. This module is where you actually *do something* with it: turn it into plots with `ggplot2`. Along the way, you'll also pick up a handful of habits (naming files, organizing scripts, recognizing R Markdown, and what version control is for) that matter for real work beyond this course, even though none of them involve writing new code.

## Basic plots with ggplot2

Every `ggplot2` plot follows the same basic grammar:

- **`ggplot(data, aes(...))`** -- start a plot, and tell it which columns to use with `aes()` (short for "aesthetic mapping") -- which column goes on the x-axis, which on the y-axis, and so on
- **`+ geom_...()`** -- add a "geometry" layer that actually draws something: `geom_point()` for a scatter plot, `geom_col()` for bars, `geom_histogram()` for a distribution

These pieces are combined with `+`, not the pipe:

```{r ggplot-demo, echo=TRUE, fig.height=3}
ggplot(fish_clean, aes(x = species, y = length_mm)) +
  geom_point()
```

```{r, echo=FALSE}
quiz_question_ui("ggplot_grammar_quiz", "What does aes() tell ggplot to do?", c(
  "Map data columns to visual properties like x, y, or color" = "correct",
  "Run the plotting code faster" = "wrong_speed",
  "Load the ggplot2 package" = "wrong_load"
))
```

### Practice: scatter plot

Below, create a scatter plot of `fish_clean` with `length_mm` on the x-axis and `weight_g` on the y-axis, replacing both `FIXME`s.

```{r, echo=FALSE}
exercise_ui("scatter_exercise", "ggplot(fish_clean, aes(x = FIXME, y = FIXME)) +\n  geom_point()", has_plot_output = TRUE)
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_1",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_1",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Bar charts from summarized data

A bar chart usually plots *summarized* data, not raw rows -- which is exactly what Module 4's `group_by()` + `summarize()` produce. You can pipe a summary straight into `ggplot()`:

```{r bar-demo, echo=TRUE, fig.height=3}
fish_clean |>
  group_by(lake_site) |>
  summarize(mean_length = mean(length_mm, na.rm = TRUE)) |>
  ggplot(aes(x = lake_site, y = mean_length)) +
  geom_col()
```

### Practice: mean weight per species

Below, group `fish_clean` by `species`, summarize each group's mean `weight_g` into `mean_weight` (already written for you), then plot the result as a bar chart with `species` on the x-axis and `mean_weight` on the y-axis. Replace all three `FIXME`s.

```{r, echo=FALSE}
exercise_ui("bar_exercise", "fish_clean |>\n  group_by(FIXME) |>\n  summarize(mean_weight = mean(weight_g, na.rm = TRUE)) |>\n  ggplot(aes(x = FIXME, y = FIXME)) +\n  geom_col()", has_plot_output = TRUE)
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_2",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_2",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Histograms

A histogram shows the *distribution* of a single numeric column -- how its values are spread out -- by grouping values into bins and counting how many fall in each one:

```{r histogram-demo, echo=TRUE, fig.height=3}
ggplot(fish_clean, aes(x = weight_g)) +
  geom_histogram(bins = 6)
```

### Practice: histogram of length

Below, create a histogram of `fish_clean`'s `length_mm` column, replacing `FIXME`.

```{r, echo=FALSE}
exercise_ui("histogram_exercise", "ggplot(fish_clean, aes(x = FIXME)) +\n  geom_histogram(bins = 8)", has_plot_output = TRUE)
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_3",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_3",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Markdown awareness

You've been reading formatted text like this the whole course -- headings, bullet points, bold text -- inside what's actually an **R Markdown** (`.Rmd`) document (this tutorial is one). R Markdown combines regular text with R code and that code's output, all in one file that renders into a finished report.

You won't be asked to author R Markdown documents in this course -- just recognize what one is and what it's for, since scripts and reports you're handed later on may well be written as one.

```{r, echo=FALSE}
quiz_question_ui("markdown_quiz", "What is an R Markdown (.Rmd) file mainly used for?", c(
  "Combining code, its output, and written text into one formatted document or report" = "correct",
  "Running R code faster than a normal script" = "wrong_speed",
  "Storing a database connection's password" = "wrong_password"
))
```

## Saving and organizing scripts

A plain R script (a `.R` file) is just your code, without any of R Markdown's surrounding text. Where you save one matters more than it might seem: keeping scripts inside your project's folder (not scattered on the Desktop or in Downloads) keeps everything a project needs in one place, so file paths -- like the working directory concepts from Module 1 -- keep working, and nothing gets lost or orphaned.

```{r, echo=FALSE}
quiz_question_ui("scripts_quiz", "Why is it better to save your R scripts inside your project folder rather than on the Desktop?", c(
  "Keeps all the files a project needs together, so paths (like working directory) still work and nothing gets lost" = "correct",
  "R scripts only run correctly if saved on the Desktop" = "wrong_required",
  "It makes the script run faster" = "wrong_speed"
))
```

## File naming conventions

A good file name describes what the file actually does, using lowercase words separated by `_` or `-`, with no spaces or special characters -- similar to this course's `snake_case` variable naming from Module 3. Avoid vague names (`asdf.R`) and especially avoid manual "versioning" in the name itself (`final_v2_FINAL.R`) -- that gets confusing fast, and it's exactly the problem real version control (next section) solves properly.

```{r, echo=FALSE}
quiz_question_ui("naming_files_quiz", "Which of these is the best file name for a script that cleans survey data?", c(
  "clean_survey_data.R" = "correct",
  "Final FINAL script (2).R" = "wrong_final",
  "asdf.R" = "wrong_vague"
))
```

## Version control awareness

**Version control** (most commonly **Git**, often paired with **GitHub** for hosting) tracks every change made to a project's files over time -- who changed what, when, and why -- so you can see a file's full history and undo a mistake instead of relying on file names like `final_v2.R` to keep track manually. This course doesn't have you use Git hands-on -- just know that it exists and roughly what it's for, since it's standard in real collaborative work.

```{r, echo=FALSE}
quiz_question_ui("version_control_quiz", "What is the main purpose of a version control system like Git?", c(
  "Tracking changes to files over time so you can see history and undo mistakes" = "correct",
  "Automatically fixing bugs in your code" = "wrong_autofix",
  "Making your R scripts run without RStudio installed" = "wrong_standalone"
))
```

## Putting it together

This last exercise combines a data-cleaning step from Module 4 with this module's plotting skills, plus one more small addition: giving a plot a title with `labs(title = "...")`.

### Practice: filtered scatter plot with a title

Below, starting from `fish_clean`: filter down to just the `"Bluegill"` rows, then create a scatter plot of `length_mm` vs. `weight_g`, and give it any title you like using `labs()`. Replace all three `FIXME`s (the third can be any text you want, in quotes).

```{r, echo=FALSE}
exercise_ui("cumulative_plot_exercise", "fish_clean |>\n  filter(FIXME) |>\n  ggplot(aes(x = length_mm, y = weight_g)) +\n  geom_point() +\n  labs(title = FIXME)", has_plot_output = TRUE)
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_4",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_4",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Wrapping up

That's the full course: from installing R and RStudio, through running and reading code, working with data, cleaning it up with dplyr, and now actually producing plots from it with ggplot2 -- plus the surrounding habits (script organization, file naming, R Markdown and version control awareness) that carry into real work beyond this tutorial.

From here, you're ready to work with the scripts, data, and instructions you'll be given in the rest of this course -- not by memorizing every function, but by recognizing what's going on and knowing how to ask for help when something doesn't make sense.
