---
title: "Module 1: Introduction to R and RStudio"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
    theme: !expr bslib::bs_theme(version = 5, bootswatch = "cerulean")
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(shiny)
library(DBI)
library(RPostgres)
library(bslib)
library(shinychat)
library(ellmer)
library(promises)

# Shared database helper functions (get_con, get_or_create_student,
# get_exercise_id, log_attempt, log_chat)
source("../R/db_utils.R")
# Tutor chat context assembly, start_tutor_chat(), and the shared
# bs5_theme_dependencies()/wire_tutor_chat()/log_attempt_and_maybe_open_chat()
# wiring used by every module's tutor chat panel.
source("../R/tutor_utils.R")

gradethis::gradethis_setup()

# Grading logic for this module's one exercise. A plain function (not a
# grade_this() object -- see module_2.Rmd's setup chunk for why) so both
# the *-check and *-error-check chunks below can share it.
evaluate_name_exercise <- function(result, stage) {
  passed <- TRUE
  feedback_message <- "Nice -- that's exactly right! Your response has been recorded."

  if (is.null(result) || !is.character(result) || nchar(trimws(result)) == 0) {
    passed <- FALSE
    feedback_message <- if (identical(stage, "error_check")) {
      "That code doesn't run as-is -- check that your name is wrapped in matching quotation marks."
    } else {
      "Make sure to type your name inside quotation marks."
    }
  } else if (identical(trimws(result), "___")) {
    passed <- FALSE
    feedback_message <- "Looks like the placeholder is still there -- replace `___` with your own name."
  }

  list(passed = passed, message = feedback_message)
}
```

```{r, context="server"}
# --- Identity capture -------------------------------------------------
# Every tutorial session starts by asking for a name. This is a
# deliberately simple stand-in for real identity -- no roster/Canvas
# matching yet. The resulting student_id is stored on session$userData,
# which is the safe, session-scoped place to keep it so that exercise
# checking code (which runs in a more isolated context) can still reach
# it via shiny::getDefaultReactiveDomain().

showModal(modalDialog(
  title = "Welcome!",
  "Please enter your name to begin this tutorial.",
  textInput("student_name", label = NULL, placeholder = "Your name"),
  footer = actionButton("submit_name", "Start Tutorial"),
  easyClose = FALSE
))

observeEvent(input$submit_name, {
  req(input$student_name)
  con <- get_con()
  sid <- get_or_create_student(con, trimws(input$student_name))
  dbDisconnect(con)

  session$userData$student_id <- sid
  removeModal()
})
```

```{r, context="server"}
# --- Tutor chat wiring --------------------------------------------------
# Always-available (collapsed by default) tutor chat panel for the one
# graded exercise. See wire_tutor_chat() in R/tutor_utils.R.

wire_tutor_chat("module1_name_check", "tutor_chat_1", "tutor_accordion_1", "Ask the Tutor", input, output, session)
```

## Welcome

```{r, echo=FALSE}
bs5_theme_dependencies()
```

This tutorial is the first step before we get into any real course material. The goal here isn't to make you a programmer -- it's to make sure you're comfortable with R and RStudio as tools, so that later material (which may come from your instructor or from an AI assistant) doesn't feel unfamiliar.

Nothing in this module involves writing real R code yet -- that starts in Module 2.

## Installing R and RStudio

R and RStudio are two different pieces of software that work together:

- **R** is the actual programming language -- the engine that runs your code.
- **RStudio** is an *interface* for working with R more comfortably -- it doesn't replace R, it sits on top of it.

You install them in that order: R first (from CRAN), then RStudio (from posit.co), since RStudio needs an existing R installation to attach to.

```{r install-quiz, echo=FALSE}
question("Which of these is true about the relationship between R and RStudio?",
  answer("RStudio replaces R -- you only need to install RStudio.", message = "Not quite -- RStudio needs R installed underneath it."),
  answer("R is the language/engine, and RStudio is an interface built on top of it.", correct = TRUE),
  answer("R and RStudio are two unrelated tools with no connection.", message = "They're closely related -- RStudio is built specifically to work with R."),
  allow_retry = TRUE
)
```

## The RStudio panes

When you open RStudio, you'll typically see four panes:

- **Source** (top-left) -- where you write and edit scripts
- **Console** (bottom-left) -- where code actually runs, one command at a time
- **Environment / History** (top-right) -- shows the objects you've created and what you've run
- **Files / Plots / Packages / Help** (bottom-right) -- a multi-purpose pane for browsing files, viewing plots, managing packages, and reading documentation

```{r panes-quiz, echo=FALSE}
question("If you want to see a plot you just created, which pane would you look at?",
  answer("Source"),
  answer("Console"),
  answer("Environment / History"),
  answer("Files / Plots / Packages / Help", correct = TRUE),
  allow_retry = TRUE
)
```

## Working directory

Your **working directory** is the folder R considers "home base" for the current session -- it's where R will look for files by default, and where it'll save things unless told otherwise.

The most reliable way to manage this is with an **RStudio Project** (a folder with a `.Rproj` file in it). Opening a project automatically sets the working directory to that project's folder, which avoids a lot of "file not found" headaches later, especially once we start importing data.

```{r wd-quiz, echo=FALSE}
question("What's the main benefit of using an RStudio Project instead of just opening loose scripts?",
  answer("It makes your code run faster.", message = "Not related to speed."),
  answer("It automatically sets your working directory to a consistent, predictable folder.", correct = TRUE),
  answer("It's required to install packages.", message = "Packages install fine either way."),
  allow_retry = TRUE
)
```

## Wrapping up

### Quick system check

Before moving on, let's confirm everything's connected properly. Replace the placeholder below with your own name, keeping the quotation marks, and click **Submit Answer**.

```{r name-exercise, exercise=TRUE}
"___"
```

```{r name-exercise-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_name_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module1_name_check", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

```{r name-exercise-error-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_name_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module1_name_check", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_1",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_1",
      placeholder = "Ask a question about this exercise...",
      greeting = "Hi! I'm here if you want to talk through this exercise. Ask me anything -- I won't just hand you the answer, but I'll help you get there.",
      allow_attachments = FALSE
    )
  )
)
```

### Up next

Module 2 introduces actual R code -- installing packages, reading the console, and running provided scripts.
