---
title: "Module 4: Clean Up"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
    theme: !expr bslib::bs_theme(version = 5, bootswatch = "cerulean")
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(DBI)
library(RPostgres)
library(bslib)
library(shinychat)
library(ellmer)
library(promises)
library(dplyr)

# Shared database helper functions (get_con, get_or_create_student,
# get_exercise_id, log_attempt, log_chat)
source("../R/db_utils.R")
# Tutor chat context assembly, start_tutor_chat(), the shared
# bs5_theme_dependencies()/wire_tutor_chat()/log_attempt_and_maybe_open_chat()
# wiring, and the quiz_question_ui()/quiz_question_server() and
# exercise_ui()/exercise_server() helpers used below in place of
# learnr::question() and exercise=TRUE/gradethis -- see their
# definitions in R/tutor_utils.R for why.
source("../R/tutor_utils.R")

# --- Practice data ---------------------------------------------------------
# A small, intentionally messy fish survey data set. Column names use the
# dot-separated, mixed-case style read.csv()/Excel exports commonly produce
# -- exactly what "Renaming columns" below has students fix. Two columns
# (Length.mm, Weight.g) include real NAs so later topics (NA handling,
# na.rm) have something genuine to work with, not a contrived edge case.
fish_survey <- data.frame(
  Fish.ID      = 1:10,
  Species.Name = c("Bluegill", "Bluegill", "Largemouth Bass", "Largemouth Bass",
                    "Yellow Perch", "Yellow Perch", "Bluegill", "Largemouth Bass",
                    "Yellow Perch", "Bluegill"),
  Length.mm    = c(152, 148, 310, NA, 205, 198, 160, 295, 210, 145),
  Weight.g     = c(45, 42, NA, 610, 98, 95, 50, 580, NA, 40),
  Lake.Site    = c("North Lake", "North Lake", "South Lake", "South Lake",
                    "North Lake", "South Lake", "North Lake", "South Lake",
                    "North Lake", "South Lake"),
  stringsAsFactors = FALSE
)

# Already-renamed version, used by every exercise AFTER the renaming one --
# so later exercises don't depend on a student having gotten renaming right.
fish_clean <- fish_survey
names(fish_clean) <- c("fish_id", "species", "length_mm", "weight_g", "lake_site")

# exercise_server() (R/tutor_utils.R) evaluates submitted code via
# eval(parse(text = user_code), envir = new.env()) -- new.env()'s parent
# chain resolves through wherever exercise_server() itself was source()'d
# (source()'s default local = FALSE means that's .GlobalEnv), NOT through
# this setup chunk's own local execution environment. Modules 1-3 never hit
# this because their exercises only ever built self-contained vectors from
# scratch -- this is the first module where exercise code needs to see a
# pre-existing shared object, so it needs to be explicitly placed in
# .GlobalEnv (safe here: both are static, read-only reference data, same
# for every session, set once when the R process starts).
assign("fish_survey", fish_survey, envir = globalenv())
assign("fish_clean", fish_clean, envir = globalenv())

# Grading logic for each exercise. log_attempt_and_maybe_open_chat() and
# wire_tutor_chat() themselves are shared across modules -- see
# R/tutor_utils.R.

evaluate_rename_exercise <- function(user_code, result, envir, stage) {
  expected_names <- c("fish_id", "species", "length_mm", "weight_g", "lake_site")
  has_fish_clean <- !is.null(envir) && exists("fish_clean", envir = envir, inherits = FALSE) &&
    is.data.frame(get("fish_clean", envir = envir))
  correct_result <- is.character(result) && identical(result, expected_names)

  passed <- has_fish_clean && correct_result

  feedback_message <- if (passed) {
    "All five columns are renamed correctly, in order -- fish_clean is ready to work with."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check that every rename() argument has the form new_name = OldName, separated by commas."
  } else if (!has_fish_clean) {
    "fish_clean doesn't exist yet -- make sure the first line assigns the renamed result to fish_clean."
  } else {
    "fish_clean exists, but the column names aren't all correct yet -- check names(fish_clean) against fish_id, species, length_mm, weight_g, lake_site, in that order."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_indexing_exercise <- function(user_code, result, stage) {
  uses_bracket <- grepl("\\[", user_code)
  correct_result <- is.data.frame(result) && nrow(result) == 5 && all(result$lake_site == "North Lake")

  passed <- uses_bracket && correct_result

  feedback_message <- if (passed) {
    "That's it -- 5 rows, all from North Lake."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check the comma inside the brackets (rows go before it, columns after -- leaving columns blank keeps them all)."
  } else if (!uses_bracket) {
    "Use square-bracket indexing here (fish_clean[ ... , ]), not a dplyr verb -- that's the point of this exercise."
  } else if (is.data.frame(result) && nrow(result) != 5) {
    "That ran, but it isn't returning exactly the 5 North Lake rows yet -- double check the comparison inside the brackets."
  } else {
    "Check that you're comparing lake_site to exactly \"North Lake\" (spelling and capitalization matter) before the comma."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_filter_select_exercise <- function(user_code, result, stage) {
  uses_filter <- grepl("filter\\s*\\(", user_code)
  uses_select <- grepl("select\\s*\\(", user_code)
  correct_result <- is.data.frame(result) &&
    identical(names(result), c("species", "length_mm")) &&
    nrow(result) == 4 &&
    all(result$species == "Bluegill")

  passed <- uses_filter && uses_select && correct_result

  feedback_message <- if (passed) {
    "Right -- 4 Bluegill rows, with just species and length_mm kept."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check the pipe (|>) connects each step, and filter()'s condition uses ==, not =."
  } else if (!uses_filter) {
    "This still needs a filter() step to keep only the Bluegill rows."
  } else if (!uses_select) {
    "This still needs a select() step to keep only species and length_mm."
  } else if (is.data.frame(result) && nrow(result) != 4) {
    "filter() and select() are both there, but the row count isn't right yet -- double check the species name inside filter()."
  } else {
    "filter() and select() both ran, but the columns kept aren't exactly species and length_mm -- check select()'s arguments."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_mutate_arrange_exercise <- function(user_code, result, stage) {
  uses_mutate <- grepl("mutate\\s*\\(", user_code)
  uses_arrange <- grepl("arrange\\s*\\(", user_code)
  has_weight_kg <- is.data.frame(result) && "weight_kg" %in% names(result)
  correct_kg <- has_weight_kg && isTRUE(all.equal(result$weight_kg, result$weight_g / 1000))
  correct_order <- is.data.frame(result) && nrow(result) > 0 && isTRUE(result$length_mm[1] == 310)

  passed <- uses_mutate && uses_arrange && correct_kg && correct_order

  feedback_message <- if (passed) {
    "weight_kg is calculated correctly, and the rows are sorted with the longest fish first."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check both mutate()'s and arrange()'s arguments are complete, valid R expressions."
  } else if (!uses_mutate || !has_weight_kg) {
    "mutate() should add a new weight_kg column, equal to weight_g divided by 1000."
  } else if (!correct_kg) {
    "weight_kg exists, but its values aren't weight_g / 1000 yet -- check the formula inside mutate()."
  } else if (!uses_arrange || !correct_order) {
    "weight_kg looks right, but the rows aren't sorted longest-to-shortest yet -- check that arrange() is using desc() on length_mm."
  } else {
    "Almost there -- double check both the mutate() and arrange() steps."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_group_summarize_exercise <- function(user_code, result, stage) {
  uses_group_by <- grepl("group_by\\s*\\(", user_code)
  uses_summarize <- grepl("summari[sz]e\\s*\\(", user_code)
  has_col <- is.data.frame(result) && "mean_length" %in% names(result)
  right_shape <- is.data.frame(result) && nrow(result) == 3
  no_na <- has_col && right_shape && !any(is.na(result$mean_length))

  passed <- uses_group_by && uses_summarize && has_col && right_shape && no_na

  feedback_message <- if (passed) {
    "Three species, each with a clean (non-NA) average length -- exactly right."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check group_by()'s and summarize()'s arguments are complete."
  } else if (!uses_group_by) {
    "This needs a group_by() step, grouping by species."
  } else if (!has_col) {
    "summarize() should create a column named exactly mean_length."
  } else if (!right_shape) {
    "Check that you're grouping by species -- there are 3 species in this data, and something else is producing a different number of groups."
  } else if (!no_na) {
    "mean_length is coming out NA for at least one species -- length_mm has a missing value, so mean() needs na.rm = TRUE."
  } else {
    "Close -- double check both steps."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_na_handling_exercise <- function(user_code, result, stage) {
  uses_filter <- grepl("filter\\s*\\(", user_code)
  uses_is_na <- grepl("is\\.na\\s*\\(", user_code)
  has_col <- is.data.frame(result) && "avg_weight" %in% names(result)
  correct_result <- has_col && !is.na(result$avg_weight[1]) && isTRUE(all.equal(result$avg_weight[1], 195))

  passed <- uses_filter && uses_is_na && correct_result

  feedback_message <- if (passed) {
    "195 -- exactly right, and the missing rows never made it into the average."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check that ! and is.na(...) are written correctly inside filter()."
  } else if (!uses_filter || !uses_is_na) {
    "filter() should keep only the rows where weight_g is NOT missing -- use is.na() with ! in front of it."
  } else if (has_col && is.na(result$avg_weight[1])) {
    "avg_weight is still coming out NA -- the missing rows aren't being filtered out yet."
  } else {
    "That's filtering something, but avg_weight isn't 195 yet -- double check the condition inside filter()."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_cumulative_exercise <- function(user_code, result, stage) {
  uses_filter <- grepl("filter\\s*\\(", user_code)
  uses_group_by <- grepl("group_by\\s*\\(", user_code)
  uses_arrange <- grepl("arrange\\s*\\(", user_code)
  right_shape <- is.data.frame(result) && nrow(result) == 2 && "avg_length" %in% names(result)
  no_na <- right_shape && !any(is.na(result$avg_length))
  right_order <- right_shape && identical(as.character(result$lake_site[1]), "South Lake")

  passed <- uses_filter && uses_group_by && uses_arrange && right_shape && no_na && right_order

  feedback_message <- if (passed) {
    "That's the full pipeline -- missing lengths filtered out, averaged per lake, and sorted with the higher average first (South Lake)."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- work through it one pipe step at a time, same as you've done in earlier modules."
  } else if (!uses_filter) {
    "Start with a filter() step that removes rows where length_mm is missing."
  } else if (!uses_group_by) {
    "This needs a group_by() step, grouping by lake_site."
  } else if (!right_shape) {
    "Check that summarize() is creating a column named exactly avg_length, and that group_by(lake_site) is producing 2 groups."
  } else if (!no_na) {
    "avg_length is coming out NA for at least one lake -- make sure the filter() step (removing missing length_mm rows) comes before the average is calculated."
  } else if (!uses_arrange || !right_order) {
    "The averages look right, but they're not sorted highest-to-lowest yet -- check that arrange() is using desc() on avg_length."
  } else {
    "Close -- double check each step of the pipeline."
  }

  list(passed = passed, message = feedback_message)
}
```

```{r, context="server"}
# --- Identity capture -------------------------------------------------
# Same pattern as module_1.Rmd/module_2.Rmd/module_3.Rmd: ask for a name on
# start, store the resulting student_id on session$userData so exercise
# check code (which runs semi-isolated from this chunk) can reach it via
# shiny::getDefaultReactiveDomain().

showModal(modalDialog(
  title = "Welcome back!",
  "Please enter your name to begin this tutorial.",
  textInput("student_name", label = NULL, placeholder = "Your name"),
  footer = actionButton("submit_name", "Start Tutorial"),
  easyClose = FALSE
))

observeEvent(input$submit_name, {
  req(input$student_name)
  tryCatch({
    con <- get_con()
    sid <- get_or_create_student(con, trimws(input$student_name))
    dbDisconnect(con)

    session$userData$student_id <- sid
    removeModal()
  }, error = function(e) {
    showModal(modalDialog(
      title = "Couldn't connect",
      paste("There was a problem saving your name (database error):", conditionMessage(e)),
      footer = actionButton("submit_name", "Try again"),
      easyClose = FALSE
    ))
  })
})
```

```{r, context="server"}
# --- Tutor chat wiring --------------------------------------------------
# Each graded exercise gets its own always-available (collapsed by
# default) tutor chat panel. See wire_tutor_chat() in R/tutor_utils.R for
# the full mechanism (fresh client per open, both directions logged).

wire_tutor_chat("module4_rename", "tutor_chat_1", "tutor_accordion_1", "Ask the Tutor", input, output, session)
wire_tutor_chat("module4_indexing", "tutor_chat_2", "tutor_accordion_2", "Ask the Tutor", input, output, session)
wire_tutor_chat("module4_filter_select", "tutor_chat_3", "tutor_accordion_3", "Ask the Tutor", input, output, session)
wire_tutor_chat("module4_mutate_arrange", "tutor_chat_4", "tutor_accordion_4", "Ask the Tutor", input, output, session)
wire_tutor_chat("module4_group_summarize", "tutor_chat_5", "tutor_accordion_5", "Ask the Tutor", input, output, session)
wire_tutor_chat("module4_na_handling", "tutor_chat_6", "tutor_accordion_6", "Ask the Tutor", input, output, session)
wire_tutor_chat("module4_cumulative", "tutor_chat_7", "tutor_accordion_7", "Ask the Tutor", input, output, session)
```

```{r, context="server"}
# --- Fix for quiz/exercise feedback outputs getting stuck "recalculating"
# if bound while their topic is hidden -- see unstick_hidden_outputs() in
# R/tutor_utils.R.
unstick_hidden_outputs(output, session)
```

```{r, context="server"}
# --- Quiz question wiring ------------------------------------------------
# See quiz_question_ui()/quiz_question_server() in R/tutor_utils.R.

quiz_question_server(input, output, session, "inspect_quiz", list(
  wrong_nrow = list(correct = FALSE, message = "nrow() only tells you the number of rows -- not each column's type."),
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_sum = list(correct = FALSE, message = "sum() only works on numeric vectors, not a whole data frame with mixed column types.")
))

quiz_question_server(input, output, session, "indexing_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_compare = list(correct = FALSE, message = "It's not comparing two data frames -- it's separating rows from columns within one."),
  wrong_rename = list(correct = FALSE, message = "Indexing brackets aren't for renaming -- that's what rename() is for.")
))

quiz_question_server(input, output, session, "pipe_quiz", list(
  correct = list(correct = TRUE, message = "Correct!"),
  wrong_multiply = list(correct = FALSE, message = "The pipe isn't arithmetic -- it just passes the left side in as the first argument to the right side."),
  wrong_dataframe_only = list(correct = FALSE, message = "The pipe works with any kind of value, not just data frames.")
))

quiz_question_server(input, output, session, "na_quiz", list(
  wrong_20 = list(correct = FALSE, message = "R doesn't just skip the NA silently -- it has to be told to."),
  correct = list(correct = TRUE, message = "Correct! Any arithmetic involving NA returns NA by default, unless you tell R how to handle it."),
  wrong_40 = list(correct = FALSE, message = "R doesn't treat a missing value as 0 -- it treats the whole result as unknown (NA).")
))
```

```{r, context="server"}
# --- Exercise wiring -------------------------------------------------------
# See exercise_ui()/exercise_server() in R/tutor_utils.R.

exercise_server(
  input, output, session, "rename_exercise", "module4_rename", "tutor_accordion_1",
  starting_code = "fish_clean <- fish_survey |> rename(\n  fish_id = FIXME,\n  species = FIXME,\n  length_mm = FIXME,\n  weight_g = FIXME,\n  lake_site = FIXME\n)\nnames(fish_clean)",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_rename_exercise(user_code, result, envir, stage)
)

exercise_server(
  input, output, session, "indexing_exercise", "module4_indexing", "tutor_accordion_2",
  starting_code = "fish_clean[fish_clean$lake_site == FIXME, ]",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_indexing_exercise(user_code, result, stage)
)

exercise_server(
  input, output, session, "filter_select_exercise", "module4_filter_select", "tutor_accordion_3",
  starting_code = "fish_clean |>\n  filter(species == FIXME) |>\n  select(species, FIXME)",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_filter_select_exercise(user_code, result, stage)
)

exercise_server(
  input, output, session, "mutate_arrange_exercise", "module4_mutate_arrange", "tutor_accordion_4",
  starting_code = "fish_clean |>\n  mutate(weight_kg = FIXME) |>\n  arrange(FIXME)",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_mutate_arrange_exercise(user_code, result, stage)
)

exercise_server(
  input, output, session, "group_summarize_exercise", "module4_group_summarize", "tutor_accordion_5",
  starting_code = "fish_clean |>\n  group_by(FIXME) |>\n  summarize(mean_length = mean(length_mm, na.rm = FIXME))",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_group_summarize_exercise(user_code, result, stage)
)

exercise_server(
  input, output, session, "na_handling_exercise", "module4_na_handling", "tutor_accordion_6",
  starting_code = "fish_clean |>\n  filter(FIXME) |>\n  summarize(avg_weight = mean(weight_g))",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_na_handling_exercise(user_code, result, stage)
)

exercise_server(
  input, output, session, "cumulative_exercise", "module4_cumulative", "tutor_accordion_7",
  starting_code = "fish_clean |>\n  filter(FIXME) |>\n  group_by(FIXME) |>\n  summarize(avg_length = mean(length_mm)) |>\n  arrange(FIXME)",
  evaluate_fn = function(user_code, result, envir, stage) evaluate_cumulative_exercise(user_code, result, stage)
)
```

## Welcome back

```{r, echo=FALSE}
bs5_theme_dependencies()
```

Modules 1-3 got you comfortable with R itself: running code, reading errors, and working with the basic building blocks (values, vectors, data frames as a concept). This module is where you start actually **working with a real data frame** -- inspecting it, cleaning it up, and asking questions of it using `dplyr`, the most common R package for this kind of work.

Throughout this module you'll use a small practice data set, `fish_survey` -- results from a (fictional) fish survey of two lakes. It looks like this:

```{r fish-survey-demo, echo=TRUE}
fish_survey
```

Real data almost never arrives this clean-looking, though -- notice the column names (`Fish.ID`, `Species.Name`, ...) and the couple of missing values (`NA`). Both of those are exactly what this module is about.

## Inspecting a data frame

Before changing anything, it's worth looking at what you actually have. A few functions for that:

- **`str(x)`** -- a compact summary: each column's name, type, and a preview of its values
- **`head(x)`** -- the first few rows
- **`nrow(x)` / `ncol(x)`** -- the number of rows / columns
- **`names(x)`** -- just the column names
- **`summary(x)`** -- basic statistics (min, max, mean, etc.) for each column

```{r inspect-demo, echo=TRUE}
str(fish_survey)
```

```{r, echo=FALSE}
quiz_question_ui("inspect_quiz", "Which function would you use to quickly see each column's data type and a preview of its values?", c(
  "nrow(fish_survey)" = "wrong_nrow",
  "str(fish_survey)" = "correct",
  "sum(fish_survey)" = "wrong_sum"
))
```

## Renaming columns

`fish_survey`'s column names (`Fish.ID`, `Species.Name`, `Length.mm`, `Weight.g`, `Lake.Site`) are a common side effect of importing data from a spreadsheet or CSV -- spaces and mixed capitalization get replaced with dots. Before doing anything else, it's worth renaming them to this course's `snake_case` convention, using dplyr's `rename()`:

```{r rename-demo, echo=TRUE}
fish_survey |> rename(species = Species.Name) |> names()
```

`rename()` always reads as `new_name = OldName` -- the new name comes first.

### Practice: rename every column

Below, rename all five columns of `fish_survey` to `fish_id`, `species`, `length_mm`, `weight_g`, and `lake_site` (in that order), replacing each `FIXME` with the matching original column name. Store the result in `fish_clean` -- you'll use that cleaned-up version for the rest of this module.

```{r, echo=FALSE}
exercise_ui("rename_exercise", "fish_clean <- fish_survey |> rename(\n  fish_id = FIXME,\n  species = FIXME,\n  length_mm = FIXME,\n  weight_g = FIXME,\n  lake_site = FIXME\n)\nnames(fish_clean)")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_1",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_1",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Indexing and subsetting

From here on, every exercise and demo uses `fish_clean` -- the already-renamed version of the data (so this module doesn't depend on your rename exercise above being correct).

Before dplyr, base R already has a way to pull out specific rows and columns, using square brackets: `df[rows, columns]`. Leaving either side blank keeps everything on that side.

```{r indexing-demo, echo=TRUE}
fish_clean[1, ]
fish_clean[, "species"]
```

You can also use a condition instead of a row number, to keep only rows matching that condition:

```{r indexing-condition-demo, echo=TRUE}
fish_clean[fish_clean$species == "Bluegill", ]
```

```{r, echo=FALSE}
quiz_question_ui("indexing_quiz", "In fish_clean[fish_clean$lake_site == \"North Lake\", ], what does the comma separate?", c(
  "Rows (before the comma) from columns (after the comma)" = "correct",
  "Two separate data frames being compared" = "wrong_compare",
  "The old column name from the new one" = "wrong_rename"
))
```

### Practice: subset by condition

Below, using square-bracket indexing (not a dplyr verb), return only the rows of `fish_clean` where `lake_site` equals `"North Lake"`, replacing `FIXME`.

```{r, echo=FALSE}
exercise_ui("indexing_exercise", "fish_clean[fish_clean$lake_site == FIXME, ]")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_2",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_2",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## The pipe operator

Bracket indexing works, but it gets hard to read once you start chaining several steps together. R's **pipe operator**, `|>`, takes the value on its left and passes it in as the first argument to the function on its right -- so `x |> mean()` is exactly the same as `mean(x)`.

```{r pipe-demo, echo=TRUE}
fish_clean$length_mm |> mean(na.rm = TRUE)
```

The real value shows up once you chain multiple steps: `x |> f() |> g()` reads left-to-right as "start with x, do f, then do g" -- which is exactly how the rest of this module's dplyr code is written.

```{r, echo=FALSE}
quiz_question_ui("pipe_quiz", "What does x |> mean() do?", c(
  "Passes x as the first argument to mean(), same as mean(x)" = "correct",
  "Multiplies x by the result of mean()" = "wrong_multiply",
  "Only works if x is a data frame" = "wrong_dataframe_only"
))
```

## Filtering and selecting

Two of dplyr's most common verbs:

- **`filter(df, condition)`** -- keep only rows matching a condition
- **`select(df, col1, col2, ...)`** -- keep only specific columns

Chained with the pipe:

```{r filter-select-demo, echo=TRUE}
fish_clean |>
  filter(lake_site == "South Lake") |>
  select(species, weight_g)
```

### Practice: filter and select

Below, starting from `fish_clean`, keep only the rows where `species` is `"Bluegill"`, then keep only the `species` and `length_mm` columns. Replace both `FIXME`s.

```{r, echo=FALSE}
exercise_ui("filter_select_exercise", "fish_clean |>\n  filter(species == FIXME) |>\n  select(species, FIXME)")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_3",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_3",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Adding columns and sorting

Two more dplyr verbs:

- **`mutate(df, new_col = expression)`** -- add a new column, calculated from existing ones
- **`arrange(df, col)`** -- sort rows by a column (ascending by default; wrap it in `desc()` for descending)

```{r mutate-arrange-demo, echo=TRUE}
fish_clean |>
  mutate(length_cm = length_mm / 10) |>
  arrange(desc(weight_g))
```

### Practice: add a column and sort

Below, starting from `fish_clean`, add a new column `weight_kg` equal to `weight_g` divided by `1000`, then sort the rows so the longest fish (by `length_mm`) comes first. Replace both `FIXME`s.

```{r, echo=FALSE}
exercise_ui("mutate_arrange_exercise", "fish_clean |>\n  mutate(weight_kg = FIXME) |>\n  arrange(FIXME)")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_4",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_4",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Grouping and summarizing

`group_by()` and `summarize()` are almost always used together: `group_by(df, col)` splits the data into groups by that column, and `summarize(df, stat = expression)` collapses each group down to one summary row.

```{r group-summarize-demo, echo=TRUE}
fish_clean |>
  group_by(lake_site) |>
  summarize(n_fish = n())
```

### Practice: average length per species

Below, group `fish_clean` by `species`, then compute each species' mean `length_mm` in a column called `mean_length`. `length_mm` has a missing value, so `mean()` will need `na.rm = TRUE` to skip it rather than returning `NA`. Replace both `FIXME`s.

```{r, echo=FALSE}
exercise_ui("group_summarize_exercise", "fish_clean |>\n  group_by(FIXME) |>\n  summarize(mean_length = mean(length_mm, na.rm = FIXME))")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_5",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_5",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Handling missing data

You just saw one way to handle `NA`s -- `na.rm = TRUE` tells a function like `mean()` to skip them. Another common approach is to remove the incomplete rows entirely first, using `filter()` with `is.na()`:

```{r, echo=FALSE}
quiz_question_ui("na_quiz", "What does mean(c(10, NA, 30)) return, without any extra arguments?", c(
  "20" = "wrong_20",
  "NA" = "correct",
  "40" = "wrong_40"
))
```

`is.na(x)` returns `TRUE` wherever `x` is missing -- so `!is.na(x)` (the `!` means "not") keeps only the rows that **aren't** missing:

```{r na-demo, echo=TRUE}
fish_clean |> filter(!is.na(weight_g))
```

### Practice: average weight, missing values removed

Below, filter `fish_clean` down to only the rows where `weight_g` is **not** missing, then compute the mean `weight_g` of the remaining rows in a column called `avg_weight`. Replace `FIXME`.

```{r, echo=FALSE}
exercise_ui("na_handling_exercise", "fish_clean |>\n  filter(FIXME) |>\n  summarize(avg_weight = mean(weight_g))")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_6",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_6",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Putting it together

You've now covered inspecting a data frame, renaming columns, indexing, the pipe, and dplyr's core verbs -- `filter()`, `select()`, `mutate()`, `arrange()`, `group_by()`, and `summarize()` -- plus handling missing values. This last exercise combines several of those into one realistic multi-step pipeline.

### Practice: full cleaning pipeline

Below, starting from `fish_clean`: filter out any row with a missing `length_mm`, group the rest by `lake_site`, compute each lake's mean `length_mm` in a column called `avg_length`, and sort the result so the lake with the **higher** average length comes first. Replace all three `FIXME`s.

```{r, echo=FALSE}
exercise_ui("cumulative_exercise", "fish_clean |>\n  filter(FIXME) |>\n  group_by(FIXME) |>\n  summarize(avg_length = mean(length_mm)) |>\n  arrange(FIXME)")
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_7",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_7",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

## Wrapping up

You've now taken a genuinely messy data frame and cleaned it up end to end: inspected it, renamed its columns, subset it, and used the pipe with `filter()`, `select()`, `mutate()`, `arrange()`, `group_by()`, and `summarize()` to answer real questions from it -- including handling missing data along the way.

### Up next

Module 5 moves into actually *doing something* with cleaned-up data: basic plots with `ggplot2`, and a light-touch introduction to a few habits (script organization, file naming, and version control awareness) that carry into real work beyond this course.
