---
title: "Module 2: Basic Code"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
    theme: !expr bslib::bs_theme(version = 5, bootswatch = "cerulean")
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(shiny)
library(DBI)
library(RPostgres)
library(bslib)
library(shinychat)
library(ellmer)
library(promises)

# Shared database helper functions (get_con, get_or_create_student,
# get_exercise_id, log_attempt, log_chat)
source("../R/db_utils.R")
# Tutor chat context assembly + start_tutor_chat()
source("../R/tutor_utils.R")

gradethis::gradethis_setup()

# One-time warm-up of learnr's exercise renderer -- see
# warm_up_exercise_renderer() in R/tutor_utils.R for why this exists.
# Runs during this setup chunk (once per app process/render), not per
# student session.
warm_up_exercise_renderer()

# Grading logic shared between each exercise's *-check and *-error-check
# chunks. These are plain functions (not grade_this() objects -- a
# grade_this() result can't safely be reused across multiple check
# chunks, since gradethis rewires its enclosing environment per call
# site) defined here in setup so both check chunks can see them; each
# *-check/*-error-check chunk is just a thin grade_this() wrapper around
# one of these. Without an explicit *-error-check chunk, a submission
# that errors (e.g. a student accidentally breaks the syntax) would
# silently fall through to gradethis's generic default error handler,
# skipping our DB logging and tutor auto-open entirely.
#
# log_attempt_and_maybe_open_chat() and wire_tutor_chat() themselves are
# shared across modules -- see R/tutor_utils.R.

evaluate_comment_exercise <- function(user_code, stage) {
  keep_active <- grepl('(?m)^\\s*print\\("Keep this line"\\)\\s*$', user_code, perl = TRUE)
  comment_present <- grepl('(?m)^\\s*#.*Comment this line out', user_code, perl = TRUE)
  comment_still_active <- grepl('(?m)^\\s*print\\("Comment this line out"\\)\\s*$', user_code, perl = TRUE)

  passed <- keep_active && comment_present && !comment_still_active

  feedback_message <- if (passed) {
    "Nice -- the first line still runs and the second is safely commented out."
  } else if (identical(stage, "error_check")) {
    "That code doesn't run as-is -- check that both lines are still valid R (matching quotes and parentheses), just with the second one commented out."
  } else if (!comment_present) {
    "The second line still isn't commented out -- add a # at the very start of that line."
  } else if (comment_still_active) {
    "Looks like there might be two copies of that line -- make sure only the commented-out version remains."
  } else {
    "Make sure the first print() line is still there and uncommented, and the second line is commented out with #."
  }

  list(passed = passed, message = feedback_message)
}

evaluate_fix_the_bug_exercise <- function(result, stage) {
  passed <- !is.null(result) && is.character(result) && identical(trimws(result), "Hello")

  feedback_message <- if (passed) {
    "Nice work -- that runs cleanly now."
  } else if (identical(stage, "error_check")) {
    "That's still erroring. Take a look at what R is telling you, and check whether Hello needs to be treated as text."
  } else {
    "The code runs without an error, but it should print the word Hello. Double check what you changed."
  }

  list(passed = passed, message = feedback_message)
}
```

```{r, context="server"}
# --- Identity capture -------------------------------------------------
# Same pattern as module_1.Rmd: ask for a name on start, store the
# resulting student_id on session$userData so exercise check code (which
# runs semi-isolated from this chunk) can reach it via
# shiny::getDefaultReactiveDomain().

showModal(modalDialog(
  title = "Welcome back!",
  "Please enter your name to begin this tutorial.",
  textInput("student_name", label = NULL, placeholder = "Your name"),
  footer = actionButton("submit_name", "Start Tutorial"),
  easyClose = FALSE
))

observeEvent(input$submit_name, {
  req(input$student_name)
  tryCatch({
    con <- get_con()
    sid <- get_or_create_student(con, trimws(input$student_name))
    dbDisconnect(con)

    session$userData$student_id <- sid
    removeModal()
  }, error = function(e) {
    showModal(modalDialog(
      title = "Couldn't connect",
      paste("There was a problem saving your name (database error):", conditionMessage(e)),
      footer = actionButton("submit_name", "Try again"),
      easyClose = FALSE
    ))
  })
})
```

```{r, context="server"}
# --- Tutor chat wiring --------------------------------------------------
# Each graded exercise gets its own always-available (collapsed by
# default) tutor chat panel. See wire_tutor_chat() in R/tutor_utils.R for
# the full mechanism (fresh client per open, both directions logged).

wire_tutor_chat("module2_comment_out", "tutor_chat_1", "tutor_accordion_1", "Ask the Tutor", input, output, session)
wire_tutor_chat("module2_fix_error", "tutor_chat_2", "tutor_accordion_2", "Ask the Tutor", input, output, session)
```

```{r, context="server"}
# --- Fix for quiz/exercise outputs getting stuck "recalculating" forever
# on any topic beyond the first -- see unstick_hidden_outputs() in
# R/tutor_utils.R for the full root-cause explanation.
unstick_hidden_outputs(output, session)
```

## Welcome back

```{r, echo=FALSE}
bs5_theme_dependencies()
```

Module 1 got R and RStudio installed and gave you a tour of the panes. This module is where you start actually touching code -- packages, the console, comments, running code someone else wrote, and (probably most usefully) how to read an error message without panicking.

You are **not** expected to write substantial code from scratch in this course. The goal is to get comfortable enough that running and adjusting code -- your instructor's, or an AI assistant's -- feels normal rather than intimidating.

## Packages

R on its own can do a lot, but most real work relies on **packages** -- bundles of extra functions other people have written. Getting a package ready to use is a two-step process:

1. **`install.packages("packagename")`** -- downloads the package onto your computer. You only need to do this once per package (until you want to update it).
2. **`library(packagename)`** -- loads that package into your *current* R session so its functions are available to use. You need to do this every time you start a new session and want to use that package.

A common beginner mix-up: running `install.packages()` again every session (unnecessary -- it's already installed) instead of `library()` (which you do need every session).

```{r packages-quiz, echo=FALSE}
question("You already installed the tidyverse package last week. You're starting a new R session today and want to use it. What do you run?",
  answer("install.packages(\"tidyverse\")", message = "That downloads it again -- not needed since it's already installed."),
  answer("library(tidyverse)", correct = TRUE),
  answer("Nothing -- it's automatically available.", message = "Not quite -- installed packages still need to be loaded with library() each session."),
  allow_retry = TRUE
)
```

## The Console

The **Console** is where code actually executes, one command at a time. When you see a prompt like `>` with nothing after it, R is idle and waiting for your next command.

Type a line of code into the Console and press Enter, and R runs it immediately -- there's no separate "run" step. This is great for quick checks (like testing what a function does), but it's not a great place to build up real work, because nothing you type there is automatically saved anywhere.

That's what **scripts** are for -- files (usually ending in `.R`) where you write code that you can save, edit, and re-run later, rather than retyping it in the Console every time.

```{r console-quiz, echo=FALSE}
question("What's the main advantage of writing your code in a script instead of typing it directly into the Console?",
  answer("Scripts run faster than the Console.", message = "Not about speed -- the code executes the same either way."),
  answer("You can save, edit, and re-run a script later; Console history is easy to lose track of.", correct = TRUE),
  answer("The Console can only run one line of code, ever.", message = "The Console can run many lines over a session -- it's just not saved as a file automatically."),
  allow_retry = TRUE
)
```

## Comments

A **comment** is a line (or part of a line) that R ignores when running your code. In R, anything after a `#` on a line is a comment.

Comments are useful for two main things:

- **Notes to yourself or others** -- explaining *why* code does something, not just what it does.
- **Temporarily turning off a line** -- if you want to skip a line without deleting it, put a `#` in front of it. This is extremely common when testing or debugging: comment a line out, see what changes, uncomment it later if you want it back.

### Practice: comment out a line

Below are two lines of code. Leave the first one exactly as it is, but use `#` to comment out the second one, so that only `"Keep this line"` gets printed when you submit. Don't delete the second line -- just comment it out.

```{r comment-exercise, exercise=TRUE}
print("Keep this line")
print("Comment this line out")
```

```{r comment-exercise-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_comment_exercise(.user_code, .stage)
  log_attempt_and_maybe_open_chat("module2_comment_out", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

```{r comment-exercise-error-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_comment_exercise(.user_code, .stage)
  log_attempt_and_maybe_open_chat("module2_comment_out", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_1",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_1",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

### Nice work

Once you've got that exercise passing, you've got the core mechanic of comments down -- everything else about them is just judgment on when to use one.

## Running code someone gives you

In this course, you'll often be handed a chunk of R code -- from your instructor, from course materials, or from an AI assistant -- and your job is to run it, not to have written it from scratch. That's a completely normal, professional way to use R, and it's the main skill this module is building.

In RStudio, the most common ways to run code are:

- **Run the current line** -- click into the line in your script and press **Ctrl+Enter** (Cmd+Return on Mac). This runs just that line in the Console and moves your cursor to the next one.
- **Run the whole script** -- click **Source** (or press Ctrl+Shift+Enter) to run every line in the file from top to bottom.

```{r running-quiz, echo=FALSE}
question("You've been given a 10-line script and just want to see what the whole thing does. What's the most efficient way to run it?",
  answer("Click into each line one at a time and press Ctrl+Enter, ten separate times.", message = "That works, but it's not the most efficient option here."),
  answer("Click Source to run the entire script at once.", correct = TRUE),
  answer("Retype all 10 lines into the Console by hand.", message = "Retyping isn't necessary -- you can run the file directly."),
  allow_retry = TRUE
)
```

## Reading error messages

Sooner or later, you'll run code and see red text instead of the output you expected. That's an **error message** -- and it's not a sign you've broken something permanently. It's R telling you, often quite specifically, what it couldn't do and why.

A few error types you'll run into constantly:

- **`object 'x' not found`** -- R looked for something named `x` and it doesn't exist yet in your session. Common cause: a typo, or forgetting to treat text as text (more on that in a second).
- **`could not find function "f"`** -- usually means a package hasn't been loaded yet with `library()`.
- **`unexpected symbol` / `unexpected ')'`** -- a syntax problem, like a missing comma, an extra or missing parenthesis, or a missing quotation mark.

One especially common mistake for beginners: writing a word without quotation marks when you meant it as text. R treats unquoted words as the *names* of objects it should already know about -- so `print(Hello)` looks for an object literally named `Hello`, not the word "Hello."

### Practice: fix the error

The code below has exactly this problem. Fix it so that it runs without an error and prints the word `Hello`.

```{r fix-the-bug-exercise, exercise=TRUE}
print(Hello)
```

```{r fix-the-bug-exercise-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_fix_the_bug_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module2_fix_error", student_id, .user_code, result$passed, result$message, "tutor_accordion_2", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

```{r fix-the-bug-exercise-error-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_fix_the_bug_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module2_fix_error", student_id, .user_code, result$passed, result$message, "tutor_accordion_2", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_2",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_2",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

### Nice work

Errors like this will keep showing up throughout the course -- and now you've got a process for them: read the message, figure out what R was actually looking for, and fix that specific thing.

## Getting help

R has built-in documentation for almost everything, and checking it is faster than it sounds:

- **`?function_name`** or **`help(function_name)`** -- opens that function's help page (arguments, examples, what it returns) in the Help pane.
- **`args(function_name)`** -- a quick shortcut that just shows you what arguments a function takes, without opening the full page.

Later in this course, you may also have an AI assistant available to ask questions like this to directly -- but `?` is still worth knowing, since it's always there, works offline, and gives you the authoritative answer straight from the package itself.

```{r help-quiz, echo=FALSE}
question("You want to know what arguments the mean() function accepts. What's the fastest way to check, from inside R?",
  answer("?mean", correct = TRUE),
  answer("Guess and see what error you get.", message = "That works eventually, but it's not the fastest or most reliable way."),
  answer("Reinstall R.", message = "Not related to checking a function's arguments."),
  allow_retry = TRUE
)
```

## Wrapping up

You've now covered packages, the console vs. scripts, comments, running provided code, reading error messages, and finding help -- the core habits that make everything after this feel far less unfamiliar.

### Up next

Module 3 gets into actual data: data types, naming conventions, vectors, assignment, and basic functions.
