---
title: "Module 1: Introduction to R and RStudio (MathJax disabled -- diagnostic variant)"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
    theme: !expr bslib::bs_theme(version = 5, bootswatch = "cerulean")
    mathjax: null
runtime: shiny_prerendered
---

<!--
DIAGNOSTIC VARIANT: identical to repro.Rmd except mathjax: null, which
should remove the <script src="https://mathjax.rstudio.com/..."> tag --
the only external (non-localhost) network request the page normally
makes. See HANDOFF.md, "New finding", for why this is being tested: an
uncaught client-side TypeError (readonly 'startTime' on an anonymous
script) appears in the app owner's real Chrome but not in automated
testing, and MathJax is the only external CDN dependency in the page.

Run with:
  rmarkdown::run("repro_no_mathjax.Rmd", shiny_args = list(port = 7413))
Then open http://127.0.0.1:7413 in the SAME real browser that shows the
bug on the normal repro.Rmd / real app, and check:
  1. Does the same "Cannot read properties of undefined (reading
     'startTime')" console error still appear?
  2. Do the quiz/exercise buttons work now?
-->

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(shiny)
library(bslib)
library(shinychat)

gradethis::gradethis_setup()

# --- STUBBED replacements for our real R/db_utils.R + R/tutor_utils.R -----
# Real project uses a Postgres DB (identity capture, attempt logging, chat
# logging) and an ellmer Anthropic chat client. None of that is needed to
# reproduce the bug, so every DB/LLM call below is a no-op/instant stub.
# Everything else (BS5 injection, accordion/chat_ui wiring shape, the
# progressive-gated exercise, multiple quiz questions across multiple
# topics) is unchanged from the real file.

TUTOR_CHAT_GREETING <- "Hi! I'm here if you want to talk through this exercise."

bs5_theme_dependencies <- function() {
  htmltools::tagList(bslib::bs_theme_dependencies(bslib::bs_theme(version = 5, bootswatch = "cerulean")))
}

get_con <- function() "stub-connection"
get_or_create_student <- function(con, display_name) "stub-student-id"
get_exercise_id <- function(con, exercise_key) 1L
log_attempt <- function(con, student_id, exercise_id, submitted_code, passed, gradethis_message) invisible(NULL)
log_chat <- function(con, student_id, exercise_id, role, message) invisible(NULL)
dbDisconnect <- function(con) invisible(NULL)

log_attempt_and_maybe_open_chat <- function(exercise_key, student_id, user_code, passed, message, accordion_id, session) {
  if (!passed) {
    bslib::accordion_panel_open(id = accordion_id, values = "Ask the Tutor", session = session)
  }
}

wire_tutor_chat <- function(exercise_key, chat_id, accordion_id, panel_value, input, output, session) {
  shiny::observeEvent(input[[accordion_id]], {
    if (panel_value %in% input[[accordion_id]]) {
      shinychat::chat_clear(chat_id, session = session)
      shinychat::chat_set_greeting(chat_id, TUTOR_CHAT_GREETING, session = session)
    }
  }, ignoreInit = TRUE, ignoreNULL = FALSE)
  # Real version also observes chat_id_user_input to stream an ellmer
  # reply; omitted here since the bug reproduces without ever sending a
  # chat message.
}

evaluate_name_exercise <- function(result, stage) {
  passed <- TRUE
  feedback_message <- "Nice -- that's exactly right! Your response has been recorded."

  if (is.null(result) || !is.character(result) || nchar(trimws(result)) == 0) {
    passed <- FALSE
    feedback_message <- if (identical(stage, "error_check")) {
      "That code doesn't run as-is -- check that your name is wrapped in matching quotation marks."
    } else {
      "Make sure to type your name inside quotation marks."
    }
  } else if (identical(trimws(result), "___")) {
    passed <- FALSE
    feedback_message <- "Looks like the placeholder is still there -- replace `___` with your own name."
  }

  list(passed = passed, message = feedback_message)
}

# --- THE ATTEMPTED FIX, verbatim from R/tutor_utils.R in the real project --
# See HANDOFF.md for the full writeup. Short version: Shiny's
# session$clientData$output_<id>_hidden flag was observed to never flip back
# to FALSE for an output bound while its containing `##` topic was hidden by
# learnr's progressive topic switching, so shouldSuspend() (which reads that
# flag) keeps such outputs permanently suspended. This function watches
# session$clientData's output_*_hidden keys on a timer and calls
# outputOptions(..., suspendWhenHidden = FALSE) on each newly-seen output
# exactly once (outputOptions() errors if called before the output exists,
# and learnr registers many quiz/exercise sub-outputs lazily, well after
# session start -- hence the polling-and-retry approach instead of a single
# upfront pass).
#
# IMPORTANT / UNRESOLVED: this appeared to fully fix the bug in Claude's own
# testing (via `Rscript R/run_tutorial.R` + Claude's own browser-automation
# tool) -- repeated, verified with direct Shiny-internals tracing
# (session$clientData dumps, shiny:bound/recalculating/recalculated/value
# event listeners, R6 Observer/ShinySession source inspection) and with real
# gradethis pass/fail feedback appearing on screen across all three modules.
# BUT: the user re-tested the exact same code via RStudio's "Run Document",
# via `Rscript R/run_tutorial.R` + their own real browser (including
# incognito, with no extensions), and reports NO change in any of those --
# the exercise/quiz buttons are still completely non-functional, exactly as
# before this fix. Debug logging below confirms the fix's own mechanism
# runs and succeeds identically in the app owner's environment too -- see
# HANDOFF.md's "New finding" section for the current leading hypothesis
# (an uncaught client-side JS error only seen in the app owner's real
# browser).
unstick_hidden_outputs <- function(output, session) {
  # TEMPORARY DIAGNOSTIC LOGGING -- confirms whether this function is even
  # running / discovering outputs / succeeding. See HANDOFF.md.
  message(
    "UNSTICK STARTED | pid=", Sys.getpid(),
    " | shiny=", getNamespaceVersion("shiny"),
    " | learnr=", getNamespaceVersion("learnr")
  )
  already_patched <- character(0)
  shiny::observe({
    shiny::invalidateLater(200, session)
    cd_names <- names(shiny::reactiveValuesToList(session$clientData))
    hidden_keys <- grep("^output_.*_hidden$", cd_names, value = TRUE)
    ids <- sub("^output_(.*)_hidden$", "\\1", hidden_keys)
    new_ids <- setdiff(ids, already_patched)
    for (id in new_ids) {
      ok <- tryCatch({
        shiny::outputOptions(output, id, suspendWhenHidden = FALSE)
        TRUE
      }, error = function(e) FALSE)
      if (ok) {
        message(
          "PATCHED: ", id,
          " | hidden=",
          session$clientData[[paste0("output_", id, "_hidden")]]
        )
        already_patched <<- c(already_patched, id)
      }
    }
  })
}
```

```{r, context="server"}
showModal(modalDialog(
  title = "Welcome!",
  "Please enter your name to begin this tutorial.",
  textInput("student_name", label = NULL, placeholder = "Your name"),
  footer = actionButton("submit_name", "Start Tutorial"),
  easyClose = FALSE
))

observeEvent(input$submit_name, {
  req(input$student_name)
  con <- get_con()
  sid <- get_or_create_student(con, trimws(input$student_name))
  dbDisconnect(con)

  session$userData$student_id <- sid
  removeModal()
})
```

```{r, context="server"}
wire_tutor_chat("module1_name_check", "tutor_chat_1", "tutor_accordion_1", "Ask the Tutor", input, output, session)
```

```{r, context="server"}
unstick_hidden_outputs(output, session)
```

## Welcome

```{r, echo=FALSE}
bs5_theme_dependencies()
```

This tutorial is the first step before we get into any real course material. The goal here isn't to make you a programmer -- it's to make sure you're comfortable with R and RStudio as tools, so that later material (which may come from your instructor or from an AI assistant) doesn't feel unfamiliar.

Nothing in this module involves writing real R code yet -- that starts in Module 2.

## Installing R and RStudio

R and RStudio are two different pieces of software that work together:

- **R** is the actual programming language -- the engine that runs your code.
- **RStudio** is an *interface* for working with R more comfortably -- it doesn't replace R, it sits on top of it.

You install them in that order: R first (from CRAN), then RStudio (from posit.co), since RStudio needs an existing R installation to attach to.

```{r install-quiz, echo=FALSE}
question("Which of these is true about the relationship between R and RStudio?",
  answer("RStudio replaces R -- you only need to install RStudio.", message = "Not quite -- RStudio needs R installed underneath it."),
  answer("R is the language/engine, and RStudio is an interface built on top of it.", correct = TRUE),
  answer("R and RStudio are two unrelated tools with no connection.", message = "They're closely related -- RStudio is built specifically to work with R."),
  allow_retry = TRUE
)
```

### Step 1: Install R

1. Go to CRAN.
2. Download and run the installer.

```{r install-r-quiz, echo=FALSE}
question("You've just downloaded the R installer file. What should you do with it?",
  answer("Delete it.", message = "No."),
  answer("Open/run it and click through the installation wizard.", correct = TRUE),
  answer("Rename it to \"RStudio\".", message = "No."),
  allow_retry = TRUE
)
```

### Step 2: Install RStudio

1. Go to posit.co.
2. Download and run the installer.

```{r install-rstudio-quiz, echo=FALSE}
question("Why install RStudio too?",
  answer("You don't need to.", message = "No."),
  answer("It's the interface on top of R.", correct = TRUE),
  answer("It installs R for you.", message = "No."),
  allow_retry = TRUE
)
```

### Step 3: Open RStudio and do a first check

Type `2 + 2` in the Console.

```{r first-check-quiz, echo=FALSE}
question("You get an error instead of [1] 4. Most likely explanation?",
  answer("RStudio is broken.", message = "No."),
  answer("R wasn't installed correctly.", correct = TRUE),
  answer("Missing math package.", message = "No."),
  allow_retry = TRUE
)
```

## The RStudio panes

Four panes: Source, Console, Environment/History, Files/Plots/Packages/Help.

```{r panes-quiz, echo=FALSE}
question("Which pane shows a plot you just created?",
  answer("Source"),
  answer("Console"),
  answer("Environment / History"),
  answer("Files / Plots / Packages / Help", correct = TRUE),
  allow_retry = TRUE
)
```

## Working directory

Use an RStudio Project.

```{r wd-quiz, echo=FALSE}
question("Main benefit of an RStudio Project?",
  answer("Faster code.", message = "No."),
  answer("Consistent working directory.", correct = TRUE),
  answer("Required for packages.", message = "No."),
  allow_retry = TRUE
)
```

## Wrapping up

### Quick system check

Replace the placeholder below with your own name, keeping the quotation marks, and click **Submit Answer**.

```{r name-exercise, exercise=TRUE}
"___"
```

```{r name-exercise-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_name_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module1_name_check", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

```{r name-exercise-error-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_name_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module1_name_check", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

Stuck, or want to double check your thinking? The tutor chat below is available any time.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_1",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_1",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

### Up next

Module 2 introduces actual R code.
