---
title: "Module 1: Introduction to R and RStudio"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
    theme: !expr bslib::bs_theme(version = 5, bootswatch = "cerulean")
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(shiny)
library(DBI)
library(RPostgres)
library(bslib)
library(shinychat)
library(ellmer)
library(promises)

# Shared database helper functions (get_con, get_or_create_student,
# get_exercise_id, log_attempt, log_chat)
source("R/db_utils.R")
# Tutor chat context assembly, start_tutor_chat(), and the shared
# bs5_theme_dependencies()/wire_tutor_chat()/log_attempt_and_maybe_open_chat()
# wiring used by every module's tutor chat panel.
source("R/tutor_utils.R")

gradethis::gradethis_setup()

# One-time warm-up of learnr's exercise renderer -- see
# warm_up_exercise_renderer() in R/tutor_utils.R for why this exists.
# Runs during this setup chunk (once per app process/render), not per
# student session.
warm_up_exercise_renderer()

# Grading logic for this module's one exercise. A plain function (not a
# grade_this() object -- see module_2.Rmd's setup chunk for why) so both
# the *-check and *-error-check chunks below can share it.
evaluate_name_exercise <- function(result, stage) {
  passed <- TRUE
  feedback_message <- "Nice -- that's exactly right! Your response has been recorded."

  if (is.null(result) || !is.character(result) || nchar(trimws(result)) == 0) {
    passed <- FALSE
    feedback_message <- if (identical(stage, "error_check")) {
      "That code doesn't run as-is -- check that your name is wrapped in matching quotation marks."
    } else {
      "Make sure to type your name inside quotation marks."
    }
  } else if (identical(trimws(result), "___")) {
    passed <- FALSE
    feedback_message <- "Looks like the placeholder is still there -- replace `___` with your own name."
  }

  list(passed = passed, message = feedback_message)
}
```

```{r, context="server"}
# --- Identity capture -------------------------------------------------
# Every tutorial session starts by asking for a name. This is a
# deliberately simple stand-in for real identity -- no roster/Canvas
# matching yet. The resulting student_id is stored on session$userData,
# which is the safe, session-scoped place to keep it so that exercise
# checking code (which runs in a more isolated context) can still reach
# it via shiny::getDefaultReactiveDomain().

showModal(modalDialog(
  title = "Welcome!",
  "Please enter your name to begin this tutorial.",
  textInput("student_name", label = NULL, placeholder = "Your name"),
  footer = actionButton("submit_name", "Start Tutorial"),
  easyClose = FALSE
))

observeEvent(input$submit_name, {
  req(input$student_name)
  tryCatch({
    con <- get_con()
    sid <- get_or_create_student(con, trimws(input$student_name))
    dbDisconnect(con)

    session$userData$student_id <- sid
    removeModal()
  }, error = function(e) {
    showModal(modalDialog(
      title = "Couldn't connect",
      paste("There was a problem saving your name (database error):", conditionMessage(e)),
      footer = actionButton("submit_name", "Try again"),
      easyClose = FALSE
    ))
  })
})
```

```{r, context="server"}
# --- Tutor chat wiring --------------------------------------------------
# Always-available (collapsed by default) tutor chat panel for the one
# graded exercise. See wire_tutor_chat() in R/tutor_utils.R.

wire_tutor_chat("module1_name_check", "tutor_chat_1", "tutor_accordion_1", "Ask the Tutor", input, output, session)
```

```{r, context="server"}
# --- Fix for quiz/exercise outputs getting stuck "recalculating" forever
# on any topic beyond the first -- see unstick_hidden_outputs() in
# R/tutor_utils.R for the full root-cause explanation.
unstick_hidden_outputs(output, session)
```

## Welcome

```{r, echo=FALSE}
bs5_theme_dependencies()
```

This tutorial is the first step before we get into any real course material. The goal here isn't to make you a programmer -- it's to make sure you're comfortable with R and RStudio as tools, so that later material (which may come from your instructor or from an AI assistant) doesn't feel unfamiliar.

Nothing in this module involves writing real R code yet -- that starts in Module 2.

## Installing R and RStudio

R and RStudio are two different pieces of software that work together:

- **R** is the actual programming language -- the engine that runs your code.
- **RStudio** is an *interface* for working with R more comfortably -- it doesn't replace R, it sits on top of it.

You install them in that order: R first (from CRAN), then RStudio (from posit.co), since RStudio needs an existing R installation to attach to.

```{r install-quiz, echo=FALSE}
question("Which of these is true about the relationship between R and RStudio?",
  answer("RStudio replaces R -- you only need to install RStudio.", message = "Not quite -- RStudio needs R installed underneath it."),
  answer("R is the language/engine, and RStudio is an interface built on top of it.", correct = TRUE),
  answer("R and RStudio are two unrelated tools with no connection.", message = "They're closely related -- RStudio is built specifically to work with R."),
  allow_retry = TRUE
)
```

This tutorial runs entirely in your browser, so it has no way to see your own computer or confirm your install actually worked -- the steps below are the real, click-by-click process, but you're doing them in a separate window, alongside this one. Keep both open side by side if you can.

### Step 1: Install R

1. Go to **[cran.r-project.org](https://cran.r-project.org)** -- this is CRAN, the official source for R. (Don't search for "R" in a general search engine first -- it's a one-letter name and the results are mostly noise. Go straight to CRAN.)
2. Near the top, click the download link for your operating system -- **Download R for Windows**, **macOS**, or **Linux**.
3. **Windows:** click **base**, then click the large **Download R-4.x.x for Windows** link. Run the downloaded `.exe` file, and click through the installer -- **Next**, **Next**, **Install**, **Finish** -- accepting every default option. You don't need to change anything.
4. **Mac:** you'll see two `.pkg` links, one for **Apple silicon (arm64)** and one for **Intel (x86_64)**. If you're not sure which chip your Mac has, check the Apple menu (top-left) > **About This Mac**. Download the matching one, open it, and follow the installer, accepting the defaults.

That's it for this step -- once the installer finishes, R itself is installed, but you won't see any new icon to click yet. That's expected; R on its own has no visible interface. That's exactly what RStudio is for, which is the next step.

```{r install-r-quiz, echo=FALSE}
question("You've just downloaded the R installer file (the .exe or .pkg). What should you do with it?",
  answer("Delete it -- you don't need it once it's downloaded.", message = "You need to actually run it to install R -- downloading alone doesn't install anything."),
  answer("Open/run it and click through the installation wizard, accepting the defaults.", correct = TRUE),
  answer("Rename it to \"RStudio\" so it opens the right program.", message = "Renaming a file doesn't change what it is -- this installs R, not RStudio, regardless of its name."),
  allow_retry = TRUE
)
```

### Step 2: Install RStudio

1. Go to **[posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/)**.
2. That page actually lists installing R as its own "Step 1" (which you already did above) -- scroll down to **Step 2: Install RStudio**.
3. Click the download button for the free **RStudio Desktop** version matching your operating system.
4. Run the installer and, same as before, click through it accepting the defaults.

```{r install-rstudio-quiz, echo=FALSE}
question("Suppose you skipped installing R and only installed RStudio. What would happen when you tried to open RStudio?",
  answer("It would work exactly the same -- RStudio doesn't actually need R.", message = "RStudio is only an interface -- it needs an actual R installation underneath it to run any code."),
  answer("It would either fail to open properly or prompt you that no R installation was found.", correct = TRUE),
  answer("It would automatically download and install R for you first.", message = "RStudio doesn't install R on its own -- the two are downloaded and installed as separate steps."),
  allow_retry = TRUE
)
```

### Step 3: Open RStudio and do a first check

Open **RStudio** now -- not R by itself; look for the RStudio icon specifically, since installing R doesn't create a separate icon for it, and R itself has no real interface to open.

The first time it launches, RStudio automatically detects the R installation from Step 1 -- you shouldn't need to point it at anything manually. You'll see a window split into a few panes (covered in detail in the next section); for now, find the **Console** pane -- it's the one with a `>` prompt waiting for input.

Click into the Console, type:

```
2 + 2
```

and press **Enter**. You should see:

```
[1] 4
```

That confirms R and RStudio are both installed and talking to each other correctly.

```{r first-check-quiz, echo=FALSE}
question("You open RStudio, type 2 + 2 in the Console, and press Enter -- but instead of [1] 4, you get an error like \"R version not found\" or the Console looks broken. What's the most likely explanation?",
  answer("RStudio is broken and needs to be reinstalled.", message = "RStudio itself is very unlikely to be the problem here -- it's almost always about what it's trying to connect to."),
  answer("R wasn't fully installed (or installed correctly) in Step 1 -- go back and reinstall R, then reopen RStudio.", correct = TRUE),
  answer("You need to install a separate \"math package\" before RStudio can do arithmetic.", message = "Basic arithmetic works in R with no extra packages -- this isn't a missing-package problem."),
  allow_retry = TRUE
)
```

## The RStudio panes

When you open RStudio, you'll typically see four panes:

- **Source** (top-left) -- where you write and edit scripts
- **Console** (bottom-left) -- where code actually runs, one command at a time
- **Environment / History** (top-right) -- shows the objects you've created and what you've run
- **Files / Plots / Packages / Help** (bottom-right) -- a multi-purpose pane for browsing files, viewing plots, managing packages, and reading documentation

```{r panes-quiz, echo=FALSE}
question("If you want to see a plot you just created, which pane would you look at?",
  answer("Source"),
  answer("Console"),
  answer("Environment / History"),
  answer("Files / Plots / Packages / Help", correct = TRUE),
  allow_retry = TRUE
)
```

## Working directory

Your **working directory** is the folder R considers "home base" for the current session -- it's where R will look for files by default, and where it'll save things unless told otherwise.

The most reliable way to manage this is with an **RStudio Project** (a folder with a `.Rproj` file in it). Opening a project automatically sets the working directory to that project's folder, which avoids a lot of "file not found" headaches later, especially once we start importing data.

```{r wd-quiz, echo=FALSE}
question("What's the main benefit of using an RStudio Project instead of just opening loose scripts?",
  answer("It makes your code run faster.", message = "Not related to speed."),
  answer("It automatically sets your working directory to a consistent, predictable folder.", correct = TRUE),
  answer("It's required to install packages.", message = "Packages install fine either way."),
  allow_retry = TRUE
)
```

## Wrapping up

### Quick system check

Before moving on, let's confirm everything's connected properly. Replace the placeholder below with your own name, keeping the quotation marks, and click **Submit Answer**.

```{r name-exercise, exercise=TRUE}
"___"
```

```{r name-exercise-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_name_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module1_name_check", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

```{r name-exercise-error-check}
grade_this({
  session <- shiny::getDefaultReactiveDomain()
  student_id <- if (!is.null(session)) session$userData$student_id else NA
  result <- evaluate_name_exercise(.result, .stage)
  log_attempt_and_maybe_open_chat("module1_name_check", student_id, .user_code, result$passed, result$message, "tutor_accordion_1", session)
  if (result$passed) pass(result$message) else fail(result$message)
})
```

Stuck, or want to double check your thinking? The tutor chat below is available any time -- before you submit, after a failed attempt, or even after you've already passed.

```{r, echo=FALSE}
accordion(
  id = "tutor_accordion_1",
  open = FALSE,
  accordion_panel(
    title = "Ask the Tutor",
    chat_ui(
      "tutor_chat_1",
      placeholder = "Ask a question about this exercise...",
      greeting = TUTOR_CHAT_GREETING,
      allow_attachments = FALSE
    )
  )
)
```

### Up next

Module 2 introduces actual R code -- installing packages, reading the console, and running provided scripts.
